---
layout: post
title: "NEON's Plant Phenology Data"
date:   2016-01-01
authors: [ ]
contributors: [ ] 
dateCreated: 2016-01-01
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [ ]
category: [self-paced-tutorial]
tags: [ ]
mainTag: 
tutorialSeries: [ ]
description: "Brief description in plain text."
code1: use-if-associated-code.R
image:
  feature: lidar_GrandMesa.png
  credit: LiDAR data collected over Grand Mesa, Colorado - National Ecological Observatory Network (NEON)
  creditlink: 
permalink: /R/plant-pheno-data
comments: true
---

{% include _toc.html %}

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

# Objectives
After completing this activity, you will:

 * 

## Things You’ll Need To Complete This Tutorial
You will need the most current version of `R` and, preferably, `RStudio` loaded
on your computer to complete this tutorial.

### Install R Packages
* **NAME:** `install.packages("NAME")`


[More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}R/Packages-In-R/)

### Download Data 
(include file that inserts a button to download data from Figshare & text about
data.  Feel free to just add text until the data set is finalized and uploaded).

{% include/dataSubsets/_data_NAME.html %}  

****
{% include/_greyBox-wd-rscript.html %}

****

## Additional Resources

* Data wrangling cheatsheet

</div>

Big Question: Plants change throughout the year - these are phenophases. 
Why do they change? 

Explore Phenology Data 

Download data  
- user guide

Briefly: 

* where is the data from (the topic of site selection)
* how is it collected 
* Which plants/growth form should we choose? 
* growth-form vs. species discussion

Final choice: - ACER


But how do we choose what aspect of phenophase to look at: 

Phenophases that are recorded (deciduous broad-leaf)/ all with Y/N & % of canopy/plant:

* breaking leaf buds 
* increasing leaf size 
* leaves 
* open flowers 
* colored leaves
* falling leaves 

Criteria for choosing 

* something with progress
* comparable across sites or growth forms? 

Final choice: leaves  (main teaching), colored leaves  (challenges)


Status codes
Status intensity - every time they go out. 
PerIndividualPerYear - once a year, "metadata" about the plants 

```{r loadStuff}

library(dplyr)
library(ggplot2)

ind <- read.csv('pheno-data/phe_perindividual.csv', stringsAsFactors = FALSE )
status <- read.csv('pheno-data/phe_statusintensity.csv', stringsAsFactors = FALSE)

```

Explore the data, let's get to know what the dataframe looks like.

```{r look-ind}
#What are the fieldnames in this dataset?
names(ind)

#how many rows are in the data?
nrow(ind)

#look at the first six rows of data.
head(ind)

# look at the structure of the dataframe.
str(ind)

```

Add protocal transect map 

Link GitHub repo - https://github.com/NEONScience/NEON-geolocation

Dates - opened first in excel (06/14/2014) vs not (2014-06-14)

```{r look-status}

# What variables are included in this dataset?
names(status)
nrow(status)
head(status)
str(status)

# min date
min(status$date)
max(status$date)

```

The `uid` is no important to understanding the data so we are going to remove uid. 
However, if you are every reporting an error in the data you should include this
with your report. 

```{r remove-uid}

ind <- select(ind,-uid)
status <- select (status, -uid)

``` 

## Clean up 

* remove duplicates (full rows)
* convert date
* retain only the latest `editedDate` in the perIndividual table.

### Remove duplicates


```{r remove-duplicates}

ind_noD <- distinct(ind)
nrow(ind_noD)

status_noD<-distinct(status)
nrow(status_noD)

```

## Figure out which names overlap 

```{r same-fieldnames}

sameName <- intersect(names(status_noD), names(ind_last))
sameName

```

Which are different: 
* editedDate
* measuredBy
* recordedBy
* samplingProtocolVersion
* remarks
* dataQF

```{r rename-column}

# rename status editedDate
status_noD <- rename(status_noD, editedDateStat=editedDate, measuredByStat=measuredBy, recordedByStat=recordedBy, samplingProtocolVersionStat=samplingProtocolVersion, remarksStat=remarks, dataQFStat=dataQF)


```


## Convert to Date

Our `addDate` and `date` columns are stored as a `character` class. We need to convert it to 
date-time class. 

```{r as-date-only }

# convert column to date class
ind_noD$editedDate <- as.Date(ind_noD$editedDate)
str(ind_noD$editedDate)

status_noD$date <- as.Date(status_noD$date)
str(status_noD$date)

```



## Retain only the latest indivdual record

Only the latest `editedDate` on ind


```{r filter-edit-date}
# retain only the max of the date for each individualID
ind_last <- ind_noD %>%
	group_by(individualID) %>%
	filter(editedDate==max(editedDate))

# oh wait, duplicate dates, retain only one
ind_lastnoD <- ind_last %>%
	group_by(editedDate, individualID) %>%
	filter(row_number()==1)

```

## Join dataframes

```{r join-dfs-error}

# Create a new dataframe "phe_ind" with all the data from status and some from ind_lastnoD
phe_ind <- left_join(status_noD, ind_lastnoD)


```

Ack!  Two different data types.  Why?  NA in taxonID is a logicial, but all the 
names are character.  

Try it again.  

`taxonID` and `scientificName` are provided for convenience in Status table, but
most up to date data is always in the `phe_perindividual.csv` files. Therefore, 
we'll remove from ...


```{r join-dfs}
# drop taxonID, scientificName
status_noD <- select (status_noD, -taxonID, -scientificName)

# Create a new dataframe "phe_ind" with all the data from status and some from ind_lastnoD
phe_ind <- left_join(status_noD, ind_lastnoD)

```


### What do we do with the data?  


Build a DF of interest with sinlgle site, species, and phenophase called `phe_1sp`.


## Select Site(s) of Interest

```{r filter-site}

# set site of interest
#subset to just the SCBI & SERC
siteOfInterest <- "SCBI"

# as you know these are the only phenophases you want to work with for now
phe_1sp <- filter(phe_ind, siteID %in% siteOfInterest)

```

## Select Species of Interest


```{r filter-species}

# see which species are present
unique(phe_AOI$taxonID)

speciesOfInterest <- "LITU"

#subset to just "LITU"
phe_1sp <- filter(phe_AOI, taxonID==speciesOfInterest)

# check that it worked
unique(phe_1sp$taxonID)

```


## Select Species of Interest

```{r filter-phonophase}

# see which species are present
unique(phe_1sp$phenophaseName)

phenophaseOfInterest <- "Leaves"

#subset to just "LITU"
phe_1sp <- filter(phe_AOI, phenophaseName==phenophaseOfInterest)

# check that it worked
unique(phe_1sp$phenophaseName)

```

## 

Calculate the phenophase total Yes of total Individuals

```{r calc-percent-yes}

```

Plot - bar graph of % of individuals in different phenophase over course of year. Interval = Daily. 


```{r}

```


Write csv (this will be read in new in subsuquent lessons)


### Set/Remove No Data Values

## Managing Missing Data: NoData values

### Find NoData Values

If we are lucky when working with external data, the `NoData` value is clearly
specified
in the metadata. No data values can be stored differently:

* **NA / nan:** Sometimes this value is `NA` or `nan` (not a number). 
* **A Designated Numeric Value (e.g. -9999):** Character strings such as `NA` can 
not always be stored along side of numeric values in some file formats. Sometimes 
you'll encounter numeric placeholders for `noData` values such as
`-9999` (a value often used in the GIS / Remote Sensing and Micrometerology 
domains.
* **Blank Values:** sometimes `noData` values are left blank. Blanks are 
particularly problematic because we can't be certain if a data value is 
purposefully missing (not measured that day or a bad measurement) or if someone 
unintentionally deleted it.

Because the actual value used to designate missing data can vary depending upon 
what data we are working with, it is important to always check the metadata for
the files associated `NoData` value. If the value is `NA`, we are in luck, `R`
will recognize and flag this value as `NoData`. If the value is numeric (e.g.,
`-9999`), then we might need to assign this value to `NA`.

<i class="fa fa-star"></i> **Data Tip:** `NA` values will be ignored when
performing calculations in `R`. However a `NoData` value of `-9999` will be
recognized as an integer and processed accordingly. If you encounter a numeric
`NoData` value be sure to assign it to `NA` in `R`:
`objectName[objectName==-9999] <- NA`
{: .notice}

In the
[Why Metadata Are Important: How to Work with Metadata in Text & EML Format]({{ site.baseurl }}/R/why-metadata-are-important/)
we viewed the metadata for these data and discovered that missing values are
designated using `NA` - a common `NoData` value placeholder. 

>**Excerpt from the metadata:** `airt: average air temperature. Average of daily 
>averages. (unit: celsius, missing value: NA)`


### Check For NoData Values  

We can quickly check for `NoData` values in our data using the`is.na()` 
function. By asking for the `sum()` of `is.na()` we can see how many NA/ missing
values we have. 

```{r missing values}

# Check for NA values
sum(is.na(phe_ind$date))
sum(is.na(phe_ind$individualID))

```

The results above tell us there are `NoData` values in the `datetime` column.
However, there are `NoData` values in other variables.  

<div id="challenge" markdown="1">

### Deal with NoData Values
When we encounter `NoData` values (blank, NaN, -9999, etc.) in our data we
need to decide how to deal with them. By default `R` treats `NoData` values
designated with a `NA` as a missing value rather than a zero. This is good, as a
value of zero (no rain today) is not the same as missing data (e.g. we didn't
measure the amount of rainfall today). 

How we deal with `NoData` values will depend on:

* the data type we are working with
* the analysis we are conducting 
* the significance of the gap or missing value

Many functions in `R` contains a `na.rm=` option which will allow you to tell R
to ignore `NA` values in your data when performing calculations.

### To Gap Fill? Or Not?

Sometimes we might need to "gap fill" our data. This means we will interpolate 
or estimate missing values often using statistical methods. Gap filling can be 
complex and is beyond the scope of this tutorial. The take away from this
is simply that it is important to acknowledge missing values in your data and to 
carefully consider how you wish to account for them during analysis. 

Other resources:

1. R code for dealing with missing data: 
<a href="http://www.statmethods.net/input/missingdata.html" target="_blank"> Quick-R: Missing Data</a> 
 
2. The Institute for Digital Research and Education has an 
<a href="http://www.ats.ucla.edu/stat/r/faq/missing.htm" target="_blank"> R FAQ on Missing Values</a>.

### Managing NoData Values in Our Data
For this tutorial, we are exploring the patterns of precipitation,
and temperature as they relate to green-up and brown-down of vegetation at 
Harvard Forest. To view overall trends during these early exploration stages, it 
is okay for us to leave out the `NoData` values in our plots. 

<i class="fa fa-star"></i> **Data Tip:** If we wanted to perform more advanced 
statistical analysis, we might consider gap-filling as our next step. Many data 
products, from towers such as FluxNet include a higher level, gap-filled
product that we can download. 
<a href="http://www.archive.arm.gov/Carbon/gapfilling/gapfilling.html"
target="_blank">More on Gap Filling</a>
{: .notice}

### NoData Values Can Impact Calculations
It is important to consider `NoData` values when performing calculations on our
data. For example, `R` will not properly calculate certain functions if there
are `NA` values in the data, unless we explicitly tell it to ignore them. 

``` {r na-in-calculations}

# calculate mean of air temperature
mean(harMet15.09.11$airt)

# are there NA values in our data?
sum(is.na(harMet15.09.11$airt))

```

`R` will not return a value for the mean as there `NA` values in the air 
temperature column. Because there are only 2 missing values (out of 105,108) for 
air temperature, we aren't that worried about a skewed 3 year mean. We can tell 
`R` to ignore noData values in the mean calculations using `na.rm=`
(NA.remove).

``` {r na-rm}
# calculate mean of air temperature, ignore NA values
mean(harMet15.09.11$airt, 
     na.rm=TRUE)

```
