---
layout: post
title: "Time & Date in Phenology Data"
date:   2016-01-01
authors: [ ]
contributors: [ ] 
dateCreated: 2016-01-01
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [ ]
category: [self-paced-tutorial]
tags: [ ]
mainTag: 
tutorialSeries: [ ]
description: "Brief description in plain text."
code1: use-if-associated-code.R
image:
  feature: lidar_GrandMesa.png
  credit: LiDAR data collected over Grand Mesa, Colorado - National Ecological Observatory Network (NEON)
  creditlink: 
permalink: 
comments: true
---

{% include _toc.html %}

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

###Goals / Objectives
After completing this activity, you will:

 * 

##Things You’ll Need To Complete This Tutorial
You will need the most current version of `R` and, preferably, `RStudio` loaded
on your computer to complete this tutorial.

###Install R Packages
* **NAME:** `install.packages("NAME")`


[More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}R/Packages-In-R/)

###Download Data 
(include file that inserts a button to download data from Figshare & text about
data.  Feel free to just add text until the data set is finalized and uploaded).

{% include/dataSubsets/_data_NAME.html %}  

****
{% include/_greyBox-wd-rscript.html %}

****

##Additional Resources

* (links if you want to provide)
* http://cflagg.github.io/codingSupportGroup/OSIS_lessons/phe_plots 

</div>

# Date & Time
## The Data Approach
In the [Intro to Time Series Data in R tutorial ]({{ site.baseurl }}/R/brief-tabular-time-series-qplot/)
we imported a time series dataset in `.csv` format into `R`. We learned how to 
quickly plot these data by converting the date column to an `R` `Date` class.
In this tutorial we will explore how to work with a column that contains both a
date AND a time stamp.

We will use functions from both base `R` and the `lubridate` package to work 
with date-time data classes.

```{r load-libraries}

# Load packages required for entire script
library(lubridate)  #work with dates

# set working directory to ensure R can find the file we wish to import
# setwd("working-dir-path-here")

```

## Import CSV File
First, let's import our time series data. We are interested in temperature, 
precipitation and photosynthetically active radiation (PAR) - metrics that are 
strongly associated with vegetation green-up and brown down (phenology or 
phenophase timing). We will use the `hf001-10-15min-m.csv` file 
that contains atmospheric data for the NEON Harvard Forest field site,
aggregated at 15-minute intervals. 

```{r read-data-csv }
# Load csv file of 15 min meteorological data from Harvard Forest
# Factors=FALSE so strings, series of letters/words/numerals, remain characters
harMet_15Min <- read.csv(
  file="NEON-DS-Met-Time-Series/HARV/FisherTower-Met/hf001-10-15min-m.csv",
  stringsAsFactors = FALSE)
```

## Date and Time Data
Let's revisit the data structure of our `harMet_15Min` object. What is the class
of the `date-time` column?

```{r view-date-structure }
# view column data class
class(harMet_15Min$datetime)

# view sample data
head(harMet_15Min$datetime)

```

Our `datetime` column is stored as a `character` class. We need to convert it to 
date-time class. What happens when we use the `as.Date` method that we learned
about in the 
[Intro to Time Series Data in R ]({{ site.baseurl }}/R/brief-tabular-time-series-qplot/ "for as.Date") 
tutorial?

```{r as-date-only }

# convert column to date class
dateOnly_HARV <- as.Date(harMet_15Min$datetime)

# view data
head(dateOnly_HARV)

```

When we use `as.Date`, we lose the time stamp. 

### Explore Date and Time Classes

#### R - Date Class - as.Date
As we just saw, the `as.Date` format doesn't store any time information. When we
use the`as.Date` method to convert a date stored as a character class to an `R`
`date` class, it will ignore all values after the date string.

```{r explore-as_date }
# Convert character data to date (no time) 
myDate <- as.Date("2015-10-19 10:15")   
str(myDate)

# what happens if the date has text at the end?
myDate2 <- as.Date("2015-10-19Hello")   
str(myDate2)

```

As we can see above the `as.Date()` function will convert the characters that it
recognizes to be part of a date into a date class and ignore all other 
characters in the string. 

#### R - Date-Time - The POSIX classes
If we have a column containing both date and time we need to use a class that
stores both date AND time. Base `R` offers two closely related classes for date
and time: `POSIXct` and `POSIXlt`. 

#### POSIXct

The `as.POSIXct` method converts a date-time string into a `POSIXct` class. 

```{r explore-POSIXct }
# Convert character data to date and time.
timeDate <- as.POSIXct("2015-10-19 10:15")   
str(timeDate)
timeDate

```

`as.POSIXct` stores both a date and time with an associated time zone. The
default time zone selected, is the time zone that your computer is set to which
is most often your local time zone. 

`POSIXct` stores date and time in seconds with the number of seconds beginning
at 1 January 1970. Negative numbers are used to store dates prior to 1970. 
Thus, the `POSIXct` format stores each date and time a single value in units of
seconds. Storing the data this way, optimizes use in `data.frames` and speeds up
computation, processing and conversion to other formats. 

```{r explore-POSIXct2 }
# to see the data in this 'raw' format, i.e., not formatted according to the 
# class type to show us a date we recognize, use the `unclass()` function.
unclass(timeDate)

```

Here we see the number of seconds from 1 January 1970 to 9 October 2015
(`1445271300`). Also, we see that a time zone (`tzone`) is associated with the value in seconds. 

<i class="fa fa-star"></i> **Data Tip:** The `unclass` method in `R` allows you
to view how a particular `R` object is stored.
{: .notice}

#### POSIXlt
The `POSIXct` format is optimized for storage and computation. However, humans 
aren't quite as computationally efficient as computers! Also, we often want to 
quickly extract some portion of the data (e.g., months). The `POSIXlt` class 
stores date and time information in a format that we are used to seeing (e.g., 
second, min, hour, day of month, month, year, numeric day of year, etc). 


```{r explore-POSIXlt }
# Convert character data to POSIXlt date and time
timeDatelt<- as.POSIXlt("2015-10-19 10:15")  
str(timeDatelt)
timeDatelt

unclass(timeDatelt)
```

When we convert a string to `POSIXlt`, and view it in `R`, it still looks
similar to the `POSIXct` format. However, `unclass()` shows us that the data are
stored differently. The `POSIXlt` class stores the hour, minute, second, day,
month, and year separately.

#### Months in POSIXlt 
`POSIXlt` has a few quirks. First, the month values stored in the `POSIXlt`
object use `zero-based indexing`. This means that month #1 (January) is stored
as 0, and month #2 (February) is stored as 1. Notice in the output above,
October is stored as the 9th month (`$mon = 9`).

#### Years in POSIXlt 
Years are also stored differently in the `POSIXlt` class. Year values are stored
using a base index value of 1900. Thus, 2015 is stored as 115 (`$year = 115` 
- 115 years since 1900).

<i class="fa fa-star"></i> **Data Tip:** To learn more about how `R` stores
information within date-time and other objects check out the `R` documentation
by using `?` (e.g., `?POSIXlt`). NOTE: you can use this same syntax to learn
about particular functions (e.g., `?ggplot`).  
{: .notice}

Having dates classified as separate components makes `POSIXlt` computationally
more resource intensive to use in `data.frames`. This slows things down! We will
thus use `POSIXct` for this tutorial.  

<i class="fa fa-star"></i> **Data Tip:** There are other `R` packages that
support date-time data classes, including `readr`, `zoo` and `chron`.  
{: .notice}

## Convert to Date-time Class

When we convert from a character to a date-time class we need to tell `R` how 
the date and time information are stored in each string. To do this, we can use
`format=`. Let's have a look at one of our date-time strings to determine it's 
format.

```{r view-date }
# view one date-time field
harMet_15Min$datetime[1]
```

Looking at the results above, we see that our data are stored in the format:
Year-Month-Day "T" Hour:Minute (`2005-01-01T00:15`). We can use this information 
to populate our `format` string using the following designations for the
components of the date-time data:

* %Y - year 
* %m - month
* %d - day
* %H:%M - hours:minutes

<i class="fa fa-star"></i> **Data Tip:** A list of options for date-time format
is available in the `strptime` function help: `help(strptime)`. Check it out!   
{: .notice }

The "T" inserted into the middle of `datetime` isn't a standard character for 
date and time, however we can tell `R` where the character is so it can ignore 
it and interpret the correct date and time as follows:
`format="%Y-%m-%dT%H:%M"`.  

```{r format-date}
# convert single instance of date/time in format year-month-day hour:min:sec
as.POSIXct(harMet_15Min$datetime[1],format="%Y-%m-%dT%H:%M")

## The format of date-time MUST match the specified format or the data will not
# convert; see what happens when you try it a different way or without the "T"
# specified
as.POSIXct(harMet_15Min$datetime[1],format="%d-%m-%Y%H:%M")
as.POSIXct(harMet_15Min$datetime[1],format="%Y-%m-%d%H:%M")

```

Using the syntax we've learned, we can convert the entire `datetime` column into 
`POSIXct` class.

```{r convert-column }
new.date.time <- as.POSIXct(harMet_15Min$datetime,
                            format="%Y-%m-%dT%H:%M" #format time
                            )

# view output
head(new.date.time)

# what class is the output
class(new.date.time)

```


### About Time Zones
Above, we successfully converted our data into a date-time class. However, what 
`timezone` is the output `new.date.time` object that we created using? 

`2005-04-15 03:30:00 MDT`

It appears as if our data were assigned MDT (which is the timezone of the
computer where these tutorials were processed originally - in Colorado - Mountain
Time). However, we know from the metadata, explored in the
[Why Metadata Are Important: How to Work with Metadata in Text & EML Format tutorial ]({{ site.baseurl }}/R/why-metadata-are-important/),
that these data were stored in Eastern Standard Time.

### Assign Time Zone

When we convert a date-time formatted column to `POSIXct` format, we need to
assign an associated **time zone**. If we don't assign a time zone,`R` will 
default to the local time zone that is defined on your computer. 
There are individual designations for different time zones and time zone 
variants (e.g., does the time occur during daylight savings time). 

<i class="fa fa-star"></i> **Data Tip:** Codes for time zones can be found in this 
<a href="https://en.wikipedia.org/wiki/List_of_tz_database_time_zones"
target="_blank">time zone table</a>. 
{: .notice}

The code for the Eastern time zone that is the closest match to the NEON Harvard
Forest field site is `America/New_York`. Let's convert our `datetime` field 
one more time, and define the associated timezone (`tz=`).

```{r assign-time-zone }
# assign time zone to just the first entry
as.POSIXct(harMet_15Min$datetime[1],
            format = "%Y-%m-%dT%H:%M",
            tz = "America/New_York")
```

The output above, shows us that the time zone is now correctly set as EST.  

### Convert to Date-time Data Class

Now, using the syntax that we learned above, we can convert the entire
`datetime` column to a `POSIXct` class.

```{r POSIX-convert-best-practice-code}
# convert to POSIXct date-time class
harMet_15Min$datetime <- as.POSIXct(harMet_15Min$datetime,
                                format = "%Y-%m-%dT%H:%M",
                                tz = "America/New_York")

# view structure and time zone of the newly defined datetime column
str(harMet_15Min$datetime)
tz(harMet_15Min$datetime)
```

Now that our `datetime` data is properly identified as a `POSIXct` date-time
data class we can continue on and look at the patterns of precipitation,
temperature, and PAR through time. 

## No Data Values


## Managing Missing Data: NoData values

### Find NoData Values

If we are lucky when working with external data, the `NoData` value is clearly
specified
in the metadata. No data values can be stored differently:

* **NA / nan:** Sometimes this value is `NA` or `nan` (not a number). 
* **A Designated Numeric Value (e.g. -9999):** Character strings such as `NA` can 
not always be stored along side of numeric values in some file formats. Sometimes 
you'll encounter numeric placeholders for `noData` values such as
`-9999` (a value often used in the GIS / Remote Sensing and Micrometerology 
domains.
* **Blank Values:** sometimes `noData` values are left blank. Blanks are 
particularly problematic because we can't be certain if a data value is 
purposefully missing (not measured that day or a bad measurement) or if someone 
unintentionally deleted it.

Because the actual value used to designate missing data can vary depending upon 
what data we are working with, it is important to always check the metadata for
the files associated `NoData` value. If the value is `NA`, we are in luck, `R`
will recognize and flag this value as `NoData`. If the value is numeric (e.g.,
`-9999`), then we might need to assign this value to `NA`.

<i class="fa fa-star"></i> **Data Tip:** `NA` values will be ignored when
performing calculations in `R`. However a `NoData` value of `-9999` will be
recognized as an integer and processed accordingly. If you encounter a numeric
`NoData` value be sure to assign it to `NA` in `R`:
`objectName[objectName==-9999] <- NA`
{: .notice}

In the
[Why Metadata Are Important: How to Work with Metadata in Text & EML Format]({{ site.baseurl }}/R/why-metadata-are-important/)
we viewed the metadata for these data and discovered that missing values are
designated using `NA` - a common `NoData` value placeholder. 

>**Excerpt from the metadata:** `airt: average air temperature. Average of daily 
>averages. (unit: celsius / missing value: NA)`

### Check For NoData Values  

We can quickly check for `NoData` values in our data using the`is.na()` 
function. By asking for the `sum()` of `is.na()` we can see how many NA/ missing
values we have. 

```{r missing values}

# Check for NA values
sum(is.na(harMet15.09.11$datetime))
sum(is.na(harMet15.09.11$airt))

# view rows where the air temperature is NA 
harMet15.09.11[is.na(harMet15.09.11$airt),]
```

The results above tell us there are `NoData` values in the `datetime` column.
However, there are `NoData` values in other variables.  

<div id="challenge" markdown="1">
## Challenge: NoData Values

How many `NoData` values are in the precipitation (`prec`) and PAR (`parr`)
columns of our data?

</div>

```{r no-data-value-challenge, echo=FALSE, results="hide"}
# check for no data values
sum(is.na(harMet15.09.11$prec))
sum(is.na(harMet15.09.11$parr))

```

### Deal with NoData Values
When we encounter `NoData` values (blank, NaN, -9999, etc.) in our data we
need to decide how to deal with them. By default `R` treats `NoData` values
designated with a `NA` as a missing value rather than a zero. This is good, as a
value of zero (no rain today) is not the same as missing data (e.g. we didn't
measure the amount of rainfall today). 

How we deal with `NoData` values will depend on:

* the data type we are working with
* the analysis we are conducting 
* the significance of the gap or missing value

Many functions in `R` contains a `na.rm=` option which will allow you to tell R
to ignore `NA` values in your data when performing calculations.

### To Gap Fill? Or Not?

Sometimes we might need to "gap fill" our data. This means we will interpolate 
or estimate missing values often using statistical methods. Gap filling can be 
complex and is beyond the scope of this tutorial. The take away from this
is simply that it is important to acknowledge missing values in your data and to 
carefully consider how you wish to account for them during analysis. 

Other resources:

1. R code for dealing with missing data: 
<a href="http://www.statmethods.net/input/missingdata.html" target="_blank"> Quick-R: Missing Data</a> 
 
2. The Institute for Digital Research and Education has an 
<a href="http://www.ats.ucla.edu/stat/r/faq/missing.htm" target="_blank"> R FAQ on Missing Values</a>.

### Managing NoData Values in Our Data
For this tutorial, we are exploring the patterns of precipitation,
and temperature as they relate to green-up and brown-down of vegetation at 
Harvard Forest. To view overall trends during these early exploration stages, it 
is okay for us to leave out the `NoData` values in our plots. 

<i class="fa fa-star"></i> **Data Tip:** If we wanted to perform more advanced 
statistical analysis, we might consider gap-filling as our next step. Many data 
products, from towers such as FluxNet include a higher level, gap-filled
product that we can download. 
<a href="http://www.archive.arm.gov/Carbon/gapfilling/gapfilling.html"
target="_blank">More on Gap Filling</a>
{: .notice}

### NoData Values Can Impact Calculations
It is important to consider `NoData` values when performing calculations on our
data. For example, `R` will not properly calculate certain functions if there
are `NA` values in the data, unless we explicitly tell it to ignore them. 

``` {r na-in-calculations}

# calculate mean of air temperature
mean(harMet15.09.11$airt)

# are there NA values in our data?
sum(is.na(harMet15.09.11$airt))

```

`R` will not return a value for the mean as there `NA` values in the air 
temperature column. Because there are only 2 missing values (out of 105,108) for 
air temperature, we aren't that worried about a skewed 3 year mean. We can tell 
`R` to ignore noData values in the mean calculations using `na.rm=`
(NA.remove).

``` {r na-rm}
# calculate mean of air temperature, ignore NA values
mean(harMet15.09.11$airt, 
     na.rm=TRUE)

```

We now see that the 3-year average air temperature is 8.5°C.  

<div id="challenge" markdown="1">
## Challenge: Import, Understand Metadata, and Clean a Data Set
We have been using the 15-minute data from the Harvard Forest. However, overall
we are interested in larger scale patterns of greening-up and browning-down.  
Thus a daily summary is sufficient for us to see overall trends.

1. Import the Daily Meteorological data from the Harvard Forest (if you haven't
already done so in the
[Intro to Time Series Data in R - Managing Date / Time Format and Simple Plots using ggplot2]({{ site.baseurl }}/R/brief-tabular-time-series-qplot/ "for HarMet_Daily")
tutorial.)
2. Check the metadata to see what the column names are for the variable of
interest (precipitation, air temperature, PAR, day and time ).
3. If needed, convert the data class of different columns.
4. Check for missing data and decide what to do with any that exist.
5. Subset out the data for the duration of our study: 2009-2011. Name the object 
"harMetDaily.09.11".
6. Export the subset to a `.csv` file. 
7. Create a plot of Daily Air Temperature for 2009-2011. Be sure to label, x-
and y-axes. Also give the plot a title! 

</div>


## Filter by phenophase of interest
Phenophase Status: Yes


## Duplicate Records
Flags: Duplicate records
Duplicate by full rows only

## Calculate the phenophase
total Yes of total Individuals

## Subset by Date
Our `.csv` file contains nearly a decade's worth of data which makes for a large
file. The time period we are interested in for our study is:

* Start Time: 1 January 2009
* End Time: 31 Dec 2011

Let's subset the data to only contain these three years. We can use the 
`subset()` function, with the syntax:
`NewObject <- subset ( ObjectToBeSubset, CriteriaForSubsetting ) `.  

We will set our criteria to be any `datetime` that:

1. Is greater than or equal to 1 Jan 2009 at 0:00 
**AND** 
2. Is less than or equal to 31 Dec 2011 at 23:59.

We also need to specify the `timezone` so `R` can handle daylight savings and
leap year.

``` {r subset-by-time }
# subset data - 2009-2011
harMet15.09.11 <- subset(harMet_15Min,
                         datetime >= as.POSIXct('2009-01-01 00:00',
                                                tz = "America/New_York") &
                         datetime <= as.POSIXct('2011-12-31 23:59',
                                               tz = "America/New_York"))

# View first and last records in the object 
head(harMet15.09.11[1])
tail(harMet15.09.11[1])

```

It worked! The first entry is 1 January 2009 at 00:00 and the last entry is 31
December 2011 at 23:45.




## Basic Plot of Phenophase 
Plot - bar graph of % of individuals in different phenophase over course of year. Interval = Daily. 
## Plot with ggplot
The `ggplot()` function within the `ggplot2` package gives us more control
over plot appearance. However, to use `ggplot` we need to learn a slightly 
different syntax. Three basic elements are needed for `ggplot()` to work:

 1. The **data_frame:** containing the variables that we wish to plot,
 2. **`aes` (aesthetics):** which denotes which variables will map to the x-, y-
 (and other) axes,  
 3. **`geom_XXXX` (geometry):** which defines the data's graphical representation
 (e.g. points (`geom_point`), bars (`geom_bar`), lines (`geom_line`), etc).
 
The syntax begins with the base statement that includes the `data_frame`
(`harMetDaily.09.11`) and associated x (`date`) and y (`airt`) variables to be
plotted:

`ggplot(harMetDaily.09.11, aes(date, airt))`

To successfully plot, the last piece that is needed is the `geom`etry type. In 
this case, we want to create a scatterplot so we can add `+ geom_point()`.

## Bar Plots with ggplot
We can use ggplot to create bar plots too. Let's create a bar plot of total 
daily precipitation next. A bar plot might be a better way to represent a total
daily value. To create a bar plot, we change the `geom` element from
`geom_point()` to `geom_bar()`.  

The default setting for a ggplot bar plot -  `geom_bar()` - is a histogram
designated by `stat="bin"`. However, in this case, we want to plot actual
precipitation values. We can use `geom_bar(stat="identity")` to force ggplot to
plot actual values.

``` {r ggplot-geom_bar }
# plot precip
PrecipDailyBarA <- ggplot(harMetDaily.09.11, aes(date, prec)) +
    geom_bar(stat="identity", na.rm = TRUE) +
    ggtitle("Daily Precipitation\n Harvard Forest") +
    xlab("Date") + ylab("Precipitation (mm)") +
    scale_x_date(labels=date_format ("%b %y"), breaks=date_breaks("1 year")) +
    theme(plot.title = element_text(lineheight=.8, face="bold", size = 20)) +
    theme(text = element_text(size=18))

PrecipDailyBarA
```

Note that some of the bars in the resulting plot appear grey rather than black.
This is because `R` will do it's best to adjust colors of bars that are closely
spaced to improve readability. If we zoom into the plot, all of the bars are
black.

<div id="challenge" markdown="1">
## Challenge: Plot with scale_x_data()
Without creating a subsetted dataframe, plot the precipitation data for 
*2010 only*.  Customize the plot with:

* a descriptive title and axis labels,
* breaks every 4 months, and
* x-axis labels as only the full month (spelled out).

HINT: you will need to rebuild the precipitation plot as you will have to 
specify a new `scale_x_data()` element. 

Bonus: Style your plot with a `ggtheme` of choice. 

</div>

``` {r ggplot-geom_bar-subset, results="hide", include=TRUE, echo=FALSE}
# Define Start and end times for the subset as R objects that are the date class
startTime2 <- as.Date("2010-01-01")
endTime2 <- as.Date("2011-01-01")

# create a start and end times R object
start.end2 <- c(startTime2,endTime2)
start.end2

# plot of precipitation
# subset just the 2011 data by using scale_x_date(limits)
ggplot(harMetDaily.09.11, aes(date, prec)) +
    geom_bar(stat="identity", na.rm = TRUE) +
    ggtitle("Daily Precipitation\n 2010\n Harvard Forest") +
    xlab("") + ylab("Precipitation (mm)") +
    scale_x_date(labels=date_format ("%B"),
    								 breaks=date_breaks("4 months"), limits=start.end2) +
    theme(plot.title = element_text(lineheight=.8, face="bold", size = 20)) +
    theme(text = element_text(size=18))

```

### Format Dates in Axis Labels
We can adjust the date display format (e.g. 2009-07 vs. Jul 09) and the number 
of major and minor ticks for axis date values using `scale_x_date`. Let's
format the axis ticks so they read "month year" (`%b %y`). To do this, we will 
use the syntax:

`scale_x_date(labels=date_format("%b %y")`

Rather than re-coding the entire plot, we can add the `scale_x_date` element
to the plot object `AirTempDaily` that we just created. 

<i class="fa fa-star"></i> **Data Tip:** You can type `?strptime` into the `R` 
console to find a list of date format conversion specifications (e.g. %b = month).
Type `scale_x_date` for a list of parameters that allow you to format dates 
on the x-axis.
{: .notice }

```{r format-x-axis-labels }
# format x-axis: dates
AirTempDailyb <- AirTempDaily + 
  (scale_x_date(labels=date_format("%b %y")))

AirTempDailyb
```

<i class="fa fa-star"></i> **Data Tip:** If you are working with a date & time
class (e.g. POSIXct), you can use `scale_x_datetime` instead of `scale_x_date`.
{: .notice }



### Export data.frame to .CSV

We can export this subset in `.csv` format to use in other analyses or to 
share with colleagues using `write.csv`. 

<i class="fa fa-star"></i> **Data Tip:** Remember, to give your output files
concise, yet descriptive names so you can identify what it contains in the
future. By default, the `.csv` file will be written to your working directory. 
{: .notice}

``` {r write-csv}
# write harMet15 subset data to .csv
write.csv(harMet15.09.11, 
          file="Met_HARV_15min_2009_2011.csv")

```

