---
layout: post
title: "Plot Continuous & Discrete Data Together"
date:   2016-01-01
authors: [ ]
contributors: [ ] 
dateCreated: 2016-01-01
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [ ]
category: [self-paced-tutorial]
tags: [ ]
mainTag: 
tutorialSeries: [ ]
description: "Brief description in plain text."
code1: use-if-associated-code.R
image:
  feature: lidar_GrandMesa.png
  credit: LiDAR data collected over Grand Mesa, Colorado - National Ecological Observatory Network (NEON)
  creditlink: 
permalink: 
comments: true
---

{% include _toc.html %}

**R Skill Level:** Intermediate - you've got the basics of `R` down.

<div id="objectives" markdown="1">

###Goals / Objectives
After completing this activity, you will:

 * 

##Things You’ll Need To Complete This Tutorial
You will need the most current version of `R` and, preferably, `RStudio` loaded
on your computer to complete this tutorial.

###Install R Packages
* **NAME:** `install.packages("NAME")`


[More on Packages in R - Adapted from Software Carpentry.]({{site.baseurl}}R/Packages-In-R/)

###Download Data 
(include file that inserts a button to download data from Figshare & text about
data.  Feel free to just add text until the data set is finalized and uploaded).

{% include/dataSubsets/_data_NAME.html %}  

****
{% include/_greyBox-wd-rscript.html %}

****

##Additional Resources

* (links if you want to provide)

</div>


Final Plot - combo of two plots stacked (e.g. not 2 y-axes on 1 plot)
Plot 1
x-axis: time, day of year
y-axis: proportion of individuals in phenophase of interest or mean % leaves
Time 
	Maybe: choose breaking leaf buds, leaves, colored leaves
Plot 2
y-axis: daily max temperature
(Maybe repeat this with multiple sites)

Science Concepts: 
Correlations between them, 
Transition periods - these are the ones to zoom in on.  

### Format Dates in Axis Labels
We can adjust the date display format (e.g. 2009-07 vs. Jul 09) and the number 
of major and minor ticks for axis date values using `scale_x_date`. Let's
format the axis ticks so they read "month year" (`%b %y`). To do this, we will 
use the syntax:

`scale_x_date(labels=date_format("%b %y")`

Rather than re-coding the entire plot, we can add the `scale_x_date` element
to the plot object `AirTempDaily` that we just created. 

<i class="fa fa-star"></i> **Data Tip:** You can type `?strptime` into the `R` 
console to find a list of date format conversion specifications (e.g. %b = month).
Type `scale_x_date` for a list of parameters that allow you to format dates 
on the x-axis.
{: .notice }

```{r format-x-axis-labels }
# format x-axis: dates
AirTempDailyb <- AirTempDaily + 
  (scale_x_date(labels=date_format("%b %y")))

AirTempDailyb
```

<i class="fa fa-star"></i> **Data Tip:** If you are working with a date & time
class (e.g. POSIXct), you can use `scale_x_datetime` instead of `scale_x_date`.
{: .notice }



### Export data.frame to .CSV

We can export this subset in `.csv` format to use in other analyses or to 
share with colleagues using `write.csv`. 

<i class="fa fa-star"></i> **Data Tip:** Remember, to give your output files
concise, yet descriptive names so you can identify what it contains in the
future. By default, the `.csv` file will be written to your working directory. 
{: .notice}

``` {r write-csv}
# write harMet15 subset data to .csv
write.csv(harMet15.09.11, 
          file="Met_HARV_15min_2009_2011.csv")

```


